---
title: "Animating plots in R"
description: |
  Using Plotly to animate plots in R.
base_url: https://mh-blog.netlify.app
preview: plot-preview.png
author:
  - name: Matthew Harris
date: 01-20-2020
output:
  distill::distill_article:
    toc: TRUE
    toc_depth: 3
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Goals

Animated visualization can be great for communicating changes over time that might
be harder to interpret from static plots. My goal is to calculate the per year 
podium counts for each of the top ten constructors. I then want to create a plot in 
Plotly that will allow me animate the change in podiums over time.

## Packages

First we need to the packages need for this analysis.

```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(plotly)
library(ggsci)
library(here)
```

## Data Import

The data used for this analysis can be found at the following link. <br>

[Formula One Data](https://github.com/mhdemo/proj_site/tree/master/content/posts/data)

```{r, include=FALSE}
read_rds("large_data/f1_master_table.rds") %>% 
  filter(!is.na(dob)) %>%
  mutate(drv_name = iconv(paste(forename, surname), 
                          from = "UTF-8", 
                          to = "ASCII", sub = ""),
         race_age = time_length(interval(dob, race_date), "years"),
         ra_group = as.character(cut(race_age, 
                                     breaks = seq(15, 60, 5),
                                     labels = c("16-20", "21-25",
                                                "26-30", "31-35",
                                                "36-40", "41-45",
                                                "46-50", "51-55", 
                                                "56-60")))) %>%
  group_by(driverId) %>%
  mutate(podium = if_else(positionOrder %in% c(1:3), 1, 0),
         win = if_else(positionOrder == 1, 1, 0)) %>%
  ungroup() -> f1_data

my_pal <- pal_npg()(7)
```

## Data Transformation

Before I can create the animated plot I need to transform the data into a format
that will be easier to work with.

```{r}
f1_data %>%
  #Groups by constructor name
  group_by(c_name) %>%
  #Calculates the total podiums by constructor
  summarize(total_p = sum(podium)) %>%
  #Filters olnly for the top 10 constructors 
  #by total podiums
  top_n(10, total_p) %>%
  #Pulls the constructor name column out as a vector
  pull(c_name) -> top_10_const

```

With the top ten constructors identified, I can compute the additional value needed for the animated plot: 
the cumulative podium wins per year. 

```{r}
f1_data %>%
  #Filters by the top constructors identified 
  #from the previous step
  filter(c_name %in% top_10_const) %>%
  #Groups by the race year and contructor name
  group_by(race_year, c_name) %>%
  #Calculates the total podiums per year for 
  #each constructor
  summarize(annual_podium = sum(podium)) %>%
  ungroup() %>%
  #Sort the data by race year to ensure that the
  #cumulative calculations work correctly
  arrange(race_year) %>%
  #Pivots the data into a wide format so that each 
  #constructor with their annual podium count is in a 
  #seperate column
  pivot_wider(names_from = c_name, 
              values_from = annual_podium) %>%
  #Replaces NA values with 0. This will ensure that the
  #cumulative podium calculation doesn't skip years where
  #there isn't data
  map_df(~if_else(is.na(.x), 0, .x)) %>%
  #Pivots the data back to the long format
  pivot_longer(cols = c(2:ncol(.)), 
               names_to = "c_name", 
               values_to = "annual_podium") %>%
  #Nest the data by constructor name
  group_by(c_name) %>%
  nest() %>%
  #Calculates the cumulative podium count for each year
  #for each consrtuctor
  mutate(cum_podiums = map(data, 
                            ~cumsum(.x$annual_podium))) %>%
  unnest(c(data, cum_podiums)) %>%
  #Drops the annual podium column
  ungroup() %>%
  select(-annual_podium) -> const_top_10

const_top_10 %>%
  head()
```

## Static plot

For comparison I have utilized our new values to visualize the podium performance with a static plot.
In order to communicate the same information I've chosen to facet the plots by year. Since there are so
many years to facet the plot becomes noisy and difficult to interpret. Now lets see how the animated plot
performs.

```{r, preview=FALSE}
const_top_10 %>%
  ggplot(aes(c_name, cum_podiums, fill = c_name)) + 
  geom_col() + facet_wrap(~race_year) + 
  theme(axis.text.x = element_text(angle = 45)) +
  scale_fill_npg() + labs(x = "Top 10 Constructors",
                          y = "Total Podiums",
                          fill = "")
```

## Creating the animated plot!!!

The animated plot is fairly easy to create. I just need to supply the
`race_year` column to the frame parameter. This will allow anyone to use the slider
to move through time and see how podiums counts have changed for each constructor. It's also way
easier to compare the performance of each constructor and identify trends.

```{r}
plot_ly(const_top_10,
        x = ~c_name,
        y = ~cum_podiums,
        color = ~c_name,
        colors = my_pal,
        type = "bar",
        frame = ~race_year) %>% 
  layout(yaxis = list(title = "Total Podiums"),
         xaxis = list(title = "Top 10 Constructors")) %>%
  animation_slider(currentvalue = list(prefix = "Year "))
```
 
